A new tidy data structure to support exploration and modeling of temporal data

### 2019-08-30 tsibble.Rmd
More minor tweaks

Perhaps, because of this variety and heterogeneity, little organization or conceptual oversight on how one should get the wild data into a tamed state is available for temporal data. Analysts are expected to do their own data preprocessing and take care of anything else needed to allow further analysis, which leads to a myriad of ad hoc solutions and duplicated efforts.

@@

Temporal data problems often fall into two types of analysis, time series and longitudinal. Both of these may have similar data input, but the representation for modelling is typically different. Time series analysis tends to focus on the dependency with series, and the cross-correlation between series. Longitudinal analysis tends to focus on overall temporal patterns across demographic or experimental treatment strata, that incorporates within subject dependency.

Time series can be univariate or multivariate, and require relatively long lengths (i.e., large $T$) for modeling. With this large $T$ property, the series can be handled as stochastic processes for the primary purpose of forecasting, and characterizing temporal dynamics. Due to an expectation of regularly spaced time, and equal lengths across series, multivariate time series are typically assumed to be in the format where each column contains a single time series, and time is specified implicitly. This also implies that data are columns of homogeneous types: either all numeric or all non-numeric. To wrestle data, in whatever format that it arrives in, to this modeling format can be frustrating. The format could be considered to be model-centric, rather than data-centric, and thus throws the analyst into the deep end of the pool, rather than allowing them to gently wade to the modeling stage from the shallow end. The expectation is that the "model" is at the center of the analytical universe. This is contrary to the **tidyverse** conceptualization (Figure \@ref(fig:tsibble-fit)), which holistically captures the data workflow. More support needs to be provided, in the form of consistent tools and data structures, to transform the data into the analytical cycle.

@@

In R [@R-base], time series and longitudinal data are of different representations. The native `ts` object and the enhancement of **zoo** [@zoo2005] and **xts** [@R-xts], assemble time series into wide matrices with implicit time indexes. If there are multiple sub-groups, like country or product type, these would be kept in different data objects. A relatively new R package **tibbletime** [@R-tibbletime] proposed a data class of *time tibble* to represent time series in heterogeneous long format. It only requires an index variable to be declared. However, this is insufficient, and a more rigid data structure is required for temporal analytics and modeling. The **plm** [@plm2008] and **panelr** [@R-panelr] packages both manage longitudinal data in long format.

@@

The choice of tidy representation of temporal data arises from a data- and model-oriented perspective, which can accommodate all of the operations that are to be performed on the data in time-based contexts. Figure \@ref(fig:tsibble-fit) marks where this new abstraction is placed in the tidy model, which is referred to as a "tsibble". Adapting the "tidy data" principles, in tsibble:

1. Index is a variable with inherent ordering from past to present.
2. Key is a set of variables that define observational units over time.
3. Each observation should be uniquely identified by index and key.
4. Each observational unit should be measured at a common interval, if regularly spaced.

Figure \@ref(fig:tsibble-semantics) sketches out the data form required for a tsibble, an extension of the tidy format to the time domain. Beyond the layout, tsibble gives the contextual meaning to variables in order to construct the temporal data object, as newly introduced "index" and "key" semantics stated in the definitions 1 and 2 above. Variables other than index and key are considered as measurements. The definitions of 3 and 4 imply that tsibble is *tidier* than tidy data, positioning itself as a model input that gives rise to more robust and reliable downstream analytics.

@@

To materialize the abstraction of the tsibble, a subset of tuberculosis cases [@tb-data], as presented in Table \@ref(tab:tb-sub), is used as an example. It contains 12 observations and 5 variables landing in a tidy data form. Each observation comprises the number of people who are diagnosed with tuberculosis for each gender at three selected countries in the years of 2011 and 2012. From tidy data to tsibble data, index and key should be declared: column `year` as the *index* variable, and column `country` together with `gender` as the *key* variable forming the observational units. Column `count` is the only measured variable in this data, but the data structure is sufficiently flexible to hold more measurements; for example, slotting the corresponding population size (if known) into the data column for normalizing the count later. Note, this data further satisfies the must-have for the distinct rows determined by index and key, and regularly spans over one-year intervals.

@@

The new tsibble structure bridges the gap between raw data and the rigorous state of temporal data analysis. The proposed contextual semantics is the new add-on to tidy data in order to support more intuitive time-related manipulations and enlighten new perspectives for time series and panel model inputs. Index, key and time interval build the three stone pillars to this new semantically structured temporal data. Each is now described in more detail.

@@

Modeling is crucial to explanatory and predictive analytics, where time series and longitudinal data analysis diverges. The tsibble, as a model-oriented object, can flow into both types of modelling, and the new semantics (index and key) can be internally utilized to accelerate modelling.

@@

To sum it up, the tsibble abstraction provides a formal organization of forwarding tidy data to model-oriented temporal data. The supporting operations can be chained for sequencing analysis, articulating a data pipeline. As @eopl stated, "No matter how complex and polished the individual operations are, it is often the quality of the glue that most directly determines the power of the system." A mini snippet below, illustrates how  transformation and forecasting are glued together, to realize the fluent pipeline.

@@

Figure \@ref(fig:animate) shows the animations of rolling windows for sliding, tiling and stretching, respectively, on annual tuberculosis cases for Australia. A block of consecutive elements with a window size of 5 are initialized and started rolling sequentially till the end of series by computing average counts.

@@

(ref:animate-cap) An illustration of window of size 5 computing rolling averages over annual tuberculosis cases in Australia with respect to sliding, tiling and stretching. (Animation needs to be viewed with Adobe Acrobat Reader.)

Rolling window adapts to functional programming, which the **purrr** package [@R-purrr] sets a good example. These functions accept and return arbitrary inputs and outputs, with arbitrary methods. For example, moving averages anticipate numerics and produce averaged numerics via `mean()`. However, rolling window regression feeds a data frame into a linear regression method like `lm()`, and generates a complex object that contains coefficients, fitted values, and etc.

Rolling window not just iterates but rolls over a sequence of elements of a fixed window. A complete and consistent set of tools are available for facilitating window-related operations, a family of `slide()`, `tile()`, `stretch()`, and their variants. `slide()` expects one input, `slide2()` two inputs, and `pslide()` multiple inputs. For type stability, the functions always return lists. Other variants including `*_lgl()`, `*_int()`, `*_dbl()`, `*_chr()` return vectors of the corresponding types, as well as `*_dfr()` and `*_dfc()` for row-binding and column-binding data frames respectively. Their multiprocessing equivalents prefixed by `future_*()` enable rolling in parallel [@R-future; @R-furrr].

@@

As a fundamental infrastructure, extensibility is a design decision that was employed from the start of **tsibble**'s development. Contrary to the "data first" principle for end users, extensibility is developer focused and would be mostly used in dependent packages, which heavily relies on S3 classes and methods in R [@adv-r]. The package can be extended in two major aspects: custom index and new tsibble class.

Time representation could be arbitrary, for example R's native `POSIXct` and `Date` for versatile date-times, nano time for nanosecond resolution in **nanotime** [@R-nanotime], and numerics in simulation. Ordered factors can also be a source of time, such as month names, January to December, and weekdays, Monday to Sunday. The **tsibble** package supports an extensive range of index types from numerics to nano time, but there might be custom indexes used for some occasions, for example school semesters. These academic terms vary from one institution to another, within an academic year which is defined differently from a calendar year. A new index would be immediately recognized upon defining `index_valid()`, as long as it can be ordered from past to future. The interval regarding semesters is further outlined through `interval_pull()`. As a result, all tsibble methods such as `has_gaps()` and `fill_gaps()` will have instant support for data that contains this new index.

@@

Sensors have been installed in households across major cities in Australia to collect data for the smart city project. One of the trials is monitoring households' electricity usage through installed smart meters in the area of Newcastle over 2010--2014 [@smart-meter]. Data from 2013 have been sliced to examine temporal patterns of customer's energy consumption with **tsibble** for this case study. Half-hourly general supply in kwH have been recorded for `r format(tsibble::n_keys(elec_ts), big.mark = ",")` customers in the data set, resulting in `r format(NROW(elec_ts), big.mark = ",")` observations in total. Daily high and low temperatures in Newcastle in 2013 provides explanatory variables other than time in a different data table [@newcastle-temp], obtained using the R package **bomrang** [@R-bomrang]. Aggregating the half-hourly energy data to the same daily time interval as the temperature data allows to join the two data tables to explore how local weather can contribute to the variations of daily electricity use and the accuracy of demand forecasting.

@@

A common practice with energy data analysis is load forecasting, because providers need to know they have capacity to supply electricity. To illustrate the pipeline including modeling, here demand is modeled for December 2013, with the usage forecast for the last day (48 steps ahead because the data is half-hourly). The energy data for the last day is not used for modeling. ARIMA models with and without a temperature covariate are fitted using automatic order selection [@autoarima]. The logarithmic transformation is applied to the average demand to ensure the positive forecasts. Figure \@ref(fig:elec-forecast) plots the one-day forecasts from both models against the actual demand, for the last two-week window. ARIMA regressed by average temperatures gives a better fit than the one without, although both tend to underestimate the night demand. The forecasting performance is reported in Table \@ref(tab:elec-accuracy), consistent with the findings in Figure \@ref(fig:elec-forecast).

>>>>

Perhaps because of this variety and heterogeneity, little organization or conceptual oversight on how one should get the wild data into a tamed state is available for temporal data. Analysts are expected to do their own data preprocessing and take care of anything else needed to allow further analysis, which leads to a myriad of ad hoc solutions and duplicated efforts.

@@

Temporal data problems often fall into two types of analysis, time series and longitudinal. Both of these may have similar data input, but the representation for modelling is typically different. Time series analysis tends to focus on the dependency within series, and the cross-correlation between series. Longitudinal analysis tends to focus on overall temporal patterns across demographic or experimental treatment strata, that incorporates within subject dependency.

Time series can be univariate or multivariate, and require relatively long lengths (i.e., large $T$) for modeling. With this large $T$ property, the series can be handled as stochastic processes for the primary purposes of forecasting and characterizing temporal dynamics. Due to an expectation of regularly spaced time, and equal lengths across series, multivariate time series are typically assumed to be in the format where each column contains a single time series, and time is specified implicitly. This also implies that data are columns of homogeneous types: either all numeric or all non-numeric. It can be frustrating to wrestle data from its original format to this modeling format. The format could be considered to be model-centric, rather than data-centric, and thus throws the analyst into the deep end of the pool, rather than allowing them to gently wade to the modeling stage from the shallow end. The expectation is that the "model" is at the center of the analytical universe. This is contrary to the **tidyverse** conceptualization (Figure \@ref(fig:tsibble-fit)), which holistically captures the data workflow. More support needs to be provided, in the form of consistent tools and data structures, to transform the data into the analytical cycle.

@@

In R [@R-base], time series and longitudinal data are of different representations. The native `ts` object and the enhancements of **zoo** [@zoo2005] and **xts** [@R-xts], assemble time series into wide matrices with implicit time indexes. If there are multiple sub-groups, such as country or product type, these would be kept in different data objects. A relatively new R package **tibbletime** [@R-tibbletime] proposed a data class of *time tibble* to represent time series in heterogeneous long format. It only requires an index variable to be declared. However, this is insufficient, and a more rigid data structure is required for temporal analytics and modeling. The **plm** [@plm2008] and **panelr** [@R-panelr] packages both manage longitudinal data in long format.

@@

The choice of tidy representation of temporal data arises from a data- and model-oriented perspective, which can accommodate all of the operations that are to be performed on the data in time-based contexts. Figure \@ref(fig:tsibble-fit) marks where this new abstraction is placed in the tidy model, which is referred to as a "tsibble". We adapt the "tidy data" principles in tsibble with the following principles:

1. The Index is a variable with inherent ordering from past to present.
2. The Key is a set of variables that define observational units over time
3. Each observation should be uniquely identified by Index and Key.
4. When each observational unit is measured at a common interval, the data set is described as "regularly spaced".

Figure \@ref(fig:tsibble-semantics) sketches out the data form required for a tsibble, an extension of the tidy format to the time domain. Beyond the layout, tsibble gives the contextual meaning to variables in order to construct the temporal data object, as newly introduced "index" and "key" semantics stated in definitions 1 and 2 above. Variables other than index and key are considered as measurements. Definitions 3 and 4 imply that a tsibble is *tidier* than tidy data, positioning itself as a model input that gives rise to more robust and reliable downstream analytics.

@@

To materialize the abstraction of the tsibble, a subset of tuberculosis cases [@tb-data], as presented in Table \@ref(tab:tb-sub), is used as an example. It contains 12 observations and 5 variables landing in a tidy data form. Each observation comprises the number of people who are diagnosed with tuberculosis for each gender at three selected countries in the years of 2011 and 2012. From tidy data to tsibble data, index and key should be declared: column `year` as the *index* variable, and column `country` together with `gender` as the *key* variables forming the observational units. Column `count` is the only measured variable in this data, but the data structure is sufficiently flexible to hold more measurements; for example, slotting the corresponding population size (if known) into the data column for normalizing the count later. Note, this data further satisfies the need for the distinct rows to be determined by index and key, and is regularly spaced over one-year intervals.

@@

The new tsibble structure bridges the gap between raw data and the rigorous state of temporal data analysis. The proposed contextual semantics is the new add-on to tidy data in order to support more intuitive time-related manipulations and enlighten new perspectives for time series and panel model inputs. The Index, Key and time interval are the three pillars to this new semantically structured temporal data. Each is now described in more detail.

@@

Modeling is crucial to explanatory and predictive analytics, where time series and longitudinal data analysis diverge. The tsibble, as a model-oriented object, can flow into both types of modelling, and the new semantics (index and key) can be internally utilized to accelerate modelling.

@@

To sum up, the tsibble abstraction provides a formal organization of forwarding tidy data to model-oriented temporal data. The supporting operations can be chained for sequencing analysis, articulating a data pipeline. As @eopl stated, "No matter how complex and polished the individual operations are, it is often the quality of the glue that most directly determines the power of the system." A mini snippet below, illustrates how  transformation and forecasting are glued together, to realize the fluent pipeline.

@@

Figure \@ref(fig:animate) shows animations of rolling windows for sliding, tiling and stretching on annual tuberculosis cases for Australia. A block of consecutive elements with a window size of 5 is initialized in each case, and the windows roll sequentially to the end of series, with average counts being computed within each window.

@@

(ref:animate-cap) An illustration of a window of size 5 to compute rolling averages over annual tuberculosis cases in Australia using sliding, tiling and stretching. (Animation needs to be viewed with Adobe Acrobat Reader.)

Rolling windows adapt to functional programming, for which the **purrr** package [@R-purrr] sets a good example. These functions accept and return arbitrary inputs and outputs, with arbitrary methods. For example, moving averages anticipate numerics and produce averaged numerics via `mean()`. However, rolling window regression feeds a data frame into a linear regression method like `lm()`, and generates a complex object that contains coefficients, fitted values, and etc.

Rolling windows not only iterate but roll over a sequence of elements of a fixed window. A complete and consistent set of tools are available for facilitating window-related operations, a family of `slide()`, `tile()`, `stretch()`, and their variants. `slide()` expects one input, `slide2()` two inputs, and `pslide()` multiple inputs. For type stability, the functions always return lists. Other variants including `*_lgl()`, `*_int()`, `*_dbl()`, `*_chr()` return vectors of the corresponding types, as well as `*_dfr()` and `*_dfc()` for row-binding and column-binding data frames respectively. Their multiprocessing equivalents prefixed by `future_*()` enable rolling in parallel [@R-future; @R-furrr].

@@

As a fundamental infrastructure, extensibility is a design decision that was employed from the start of **tsibble**'s development. Contrary to the "data first" principle for end users, extensibility is developer focused and would be mostly used in dependent packages; it heavily relies on S3 classes and methods in R [@adv-r]. The package can be extended in two major ways: custom indexes and new tsibble classes.

Time representation could be arbitrary, for example R's native `POSIXct` and `Date` for versatile date-times, nano time for nanosecond resolution in **nanotime** [@R-nanotime], and numerics in simulation. Ordered factors can also be a source of time, such as month names, January to December, and weekdays, Monday to Sunday. The **tsibble** package supports an extensive range of index types from numerics to nano time, but there might be custom indexes used for some occasions, for example school semesters. These academic terms vary from one institution to another, with the academic year defined differently from a calendar year. A new index would be immediately recognized upon defining `index_valid()`, as long as it can be ordered from past to future. The interval regarding semesters is further outlined through `interval_pull()`. As a result, all tsibble methods such as `has_gaps()` and `fill_gaps()` will have instant support for data that contains this new index.

@@

Sensors have been installed in households across major cities in Australia to collect data for the smart city project. One of the trials is monitoring households' electricity usage through installed smart meters in the area of Newcastle over 2010--2014 [@smart-meter]. Data from 2013 have been sliced to examine temporal patterns of customers' energy consumption with **tsibble** for this case study. Half-hourly general supply in kwH have been recorded for `r format(tsibble::n_keys(elec_ts), big.mark = ",")` customers in the data set, resulting in `r format(NROW(elec_ts), big.mark = ",")` observations in total. Daily high and low temperatures in Newcastle in 2013 provide explanatory variables other than time in a different data table [@newcastle-temp], obtained using the R package **bomrang** [@R-bomrang]. Aggregating the half-hourly energy data to the same daily time interval as the temperature data allows us to join the two data tables to explore how local weather can contribute to the variations of daily electricity use and the accuracy of demand forecasting.

@@

A common practice with energy data analysis is load forecasting, because providers need to know they have capacity to supply electricity. To illustrate the pipeline including modeling, we model demand for December 2013, with the usage forecast for the last day (48 steps ahead because the data is half-hourly). The energy data for the last day is not used for modeling. ARIMA models with and without a temperature covariate are fitted using automatic order selection [@autoarima]. The logarithmic transformation is applied to the average demand to ensure positive forecasts. Figure \@ref(fig:elec-forecast) plots one-day forecasts from both models against the actual demand, for the last two-week window. The ARIMA model which includes the average temperature covariate gives a better fit than the one without, although both tend to underestimate the night demand. The forecasting performance is reported in Table \@ref(tab:elec-accuracy), consistent with the findings in Figure \@ref(fig:elec-forecast).
